Online Verification of Commutativity

### 2020-09-10 acmPaper/paper.tex
small change

We then create a set $P$ of all the paths in the diagram with no cycles, and filter the set $P \times P$, excluding pairs where the paths begin or end on different nodes, or are identical, to get the set of all cycle-free parallel pairs $Q$.
After verifying $C$, it is sufficient to verify only $Q$ (as opposed to all pairs) because cycles must now be the identity, so for any pair in the set of all parallel paths, any instance of a cycle can be removed to obtain an equivalent pair with shorter, cycle-free paths.

@@

This narrowing can be done because the original diagram commutes.
Pairs where both paths do not involve the new edge would remain equal (this would apply to cycles too; cycles that do not pass through the new edge must be the identity).
Also, pairs where both paths involved the new edge would have to be equal.
To see why this is true, each path could be thought of as consisting of the composition of three segments.
For path pair $p$, and new edge from node $S$ to node $T$, the first segment extends from $\partial(p)^+$ to the $S$, the second, the new edge $(S, T)$ itself, and the third, from $T$ to $\partial(p)^-$.
The new edge could only appear once because cycles have already been dealt with so only pairs where the path includes the new edge once need be checked.
The first segment of both pairs would have to be equal because they existed as parallel pairs in the original diagram, and similarly the third segments would also have to be equal.
The second segments, consisting of the same edge, would also have to be equal because the equivalence oracle is reflexive.
Therefore, the two paths of the pair, a composition of these three equal components, would be equal, since the oracle would preserve transitivity of equality.
We are left only with parallel pairs where exactly one of the paths passes through the new edge.

@@

We present a polynomial time solution to the \textsc{verification set problem}.
As in the online baseline algorithm, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation allowing us to improve on the online baseline is a result of Theorem~\ref{reductionRule} (which we expand on later): for a given source and sink pair, only a single parallel pair needs to be verified.
Theorem~\ref{verifyingSet} shows that, should our selected set of pairs and cycles passing through the new edge be verified commutative, the entire diagram must commute.
\ref{algo_online_polynomial} uses this strategy of identifying a parallel pair with exactly one path through the edge for each (source, sink) pair.

\begin{algorithm}[t]
\DontPrintSemicolon
\KwData{existing graph, new edge.}
\KwResult{Set of parallel pairs to verify.}
\SetKwProg{try}{try}{:}{}
\SetKwProg{catch}{catch}{:}{end}
Graph existingGraph;
Edge newEdge;

@@

The \code{try} block is executed at most $O(|V|^2)$ times, which is also the bound on the number of pairs verified.
%
This bound is asymptotically tight, as can be seen in the case where the graph contains $2N$ nodes along $S$ and $T$.
Imagine dividing the nodes into two groups of $N$ nodes each. Every node in group 1 has a forward edge to every node in group 2 and to $S$. $T$ has a forward edge to every node in group 2. In this diagram, when adding edge $(S, T)$, $N^2$ paths need to be verified which is polynomial in the total number of nodes, $2N+2$.

@@

\begin{theorem}
\label{reductionRule}
If parallel paths $g_2 = f_2; n; h_2$ then it must be that $g_1 = f_1; n; h_1$.
\end{theorem} 
% monomorphisms
\begin{proof}
We use the fact that $f_1$=$l; f_2$ and $h_1$=$h_2; m$.
\[g_2 = f_2; n; h_2 \Rightarrow l; g_2 = l; f_2; n ; h_2 \]
\[\Rightarrow l ; g_2 ; m = l ; f_2 ; n ; h_2 ; m \Rightarrow g_1 = f_1 ; n ; h_1\]
%
The proof holds if any of the paths used are the identity, e.g., if $f_1$ is the identity so $S$ and $P_1$ are the same node.
\end{proof}

@@

Given that a set of path pairs are equal, suppose we attempt to derive the proposition that a different parallel pair of paths is equal with a step-by-step application of inference rules.
 Under the assumption that edges are generic functions, and no other information is available, $F$ is a semi-group.
 The only inference rules allowed are composition (given that $f_1=f_2$, it must be that $g ; f_1 = g ; f_2$) and replacement of one path by a different, equal path (given $f_1=f_2$ and $g ; f_1 = h ; f_1 $, it must be true that $g ; f_1 = h ; f_2$). Any permutation of the repeated application of these two rules results in the ``reduction rule'' already described; it is therefore the only rule that can be used to reduce the set of path pairs to check.

@@

The na\"{i}ve baseline performs poorly, taking well over a thousand seconds for even small graphs of 10 nodes.
While the batch algorithm improves on this, it still does not scale very well, with computation for a graph with 14 nodes and 0.4 density taking hours.
Our implementation does not memoize the construction of the vector and matrix representation of paths in GF(2); profiling indicates that this construction is a major factor in the high time cost for these algorithms.
\ref{algo_online_polynomial} performs only slightly better than the batch algorithm.
Surprisingly, the optimal set algorithm cuts time cost by several orders of magnitude, and runs in milliseconds for small graphs.
All implementations are sensitive to density, performing better when density is low.

@@

>>>>

We then create a set $P$ of all the paths in the diagram with no cycles, and filter the set $P \times P$, excluding pairs where the paths begin or end on different nodes, or are identical, to get the set of all cycle-free parallel pairs $Q$.
After verifying $C$, it is sufficient to verify only $Q$ (as opposed to all pairs, per Lemma~\ref{one_occurence_lemma}) because cycles must now be the identity, so for any pair in the set of all parallel paths, any instance of a cycle can be removed to obtain an equivalent pair with shorter, cycle-free paths.

@@

This narrowing can be done because the original diagram commutes.
Pairs where both paths do not involve the new edge would remain equal (this would apply to cycles too; cycles that do not pass through the new edge must be the identity).
Also, pairs where both paths involved the new edge would have to be equal.
To see why this is true, each path could be thought of as consisting of the composition of three segments.
For path pair $p$, and new edge from node $S$ to node $T$, the first segment extends from $\partial(p)^+$ to the $S$, the second, the new edge $(S, T)$ itself, and the third, from $T$ to $\partial(p)^-$.
The new edge could only appear once because cycles have already been dealt with so only pairs where the path includes the new edge once need be checked (Lemma~\ref{one_occurence_lemma}).
The first segment of both pairs would have to be equal because they existed as parallel pairs in the original diagram, and similarly the third segments would also have to be equal.
The second segments, consisting of the same edge, would also have to be equal because the equivalence oracle is reflexive.
Therefore, the two paths of the pair, a composition of these three equal components, would be equal, since the oracle would preserve transitivity of equality.
We are left only with parallel pairs where exactly one of the paths passes through the new edge.

@@

We present a polynomial time solution to the \textsc{verification set problem} .
As in the online baseline algorithm, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation allowing us to improve on the online baseline is a result of Theorem~\ref{reductionRule} (which will be expanded on later): for a given source and sink pair, only a single parallel pair needs to be verified.
Theorem~\ref{verifyingSet} shows that, should our selected set of pairs and cycles passing through the new edge be verified commutative, the entire diagram must commute.
Algorithm \ref{algo_online_polynomial} uses this strategy of identifying a parallel pair with exactly one path through the edge for each (source, sink) pair.

\begin{algorithm}
\DontPrintSemicolon
\KwData{existing graph, new edge.}
\KwResult{Set of parallel pairs to verify.}
\SetKwProg{try}{try}{:}{}
\SetKwProg{catch}{catch}{:}{end}
Graph existingGraph;
Edge newEdge;

@@

The try block is executed at most $O(|V|^2)$ times, which is also the bound on the number of pairs verified.
%
This bound is asymptotically tight, as can be seen in the case where the graph contains $2N$ nodes along $S$ and $T$.
We imagine dividing the nodes into two groups of $N$ nodes each. Every node in group 1 has a forward node to every node in group 2 as well as to $S$. $T$ has a forward edge to every node in group 2. In this diagram, on the addition of edge $(S, T)$, $N^2$ paths need to be verified which is polynomial in the total number of nodes, $2N+2$.

@@

\begin{theorem}
\label{reductionRule}
If parallel paths $g_2 = f_2; n; h_2$ then it must be that $g_1 = f_1; n; h_1$.
\end{theorem} 
% monomorphisms
\begin{proof}
We use the fact that $f_1$=$l; f_2$ and $h_1$=$h_2; m$.
\[g_2 = f_2; n; h_2 \Rightarrow l; g_2 = l; f_2; n ; h_2 \]
\[\Rightarrow l ; g_2 ; m = l ; f_2 ; n ; h_2 ; m \Rightarrow g_1 = f_1 ; n ; h_1\]

The proof is not affected if any of the paths used are the identity, e.g., if $f_1$ is the identity and $S$ and $P_1$ are the same node.
\end{proof}

@@

 Given that a set of path pairs are equal, suppose that one attempts to derive the proposition that a different parallel pair of paths is equal with a step-by-step application of inference rules.
 Under the assumption that edges are generic functions, and no other information is available, $F$ is a semi-group.
 The only inference rules allowed are composition (given that $f_1=f_2$, it must be that $g ; f_1 = g ; f_2$) and replacement of one path by a different, equal path (given $f_1=f_2$ and $g ; f_1 = h ; f_1 $, it must be true that $g ; f_1 = h ; f_2$). Any permutation of the repeated application of these two rules results in the ``reduction rule'' already described; it is therefore the only rule that can be used to reduce the set of path pairs to check.

@@

The na\"{i}ve baseline performs poorly, taking well over a thousand seconds for even small graphs of 10 nodes.
While the batch algorithm improves on this, it still does not scale very well, with computation for a graph with 14 nodes and 0.4 density taking hours.
Our implementation does not memoize the construction of the vector and matrix representation of paths in GF(2); profiling indicates that this construction is a major factor in the high time cost for this algorithm.
\ref{algo_online_polynomial} performs only slightly better than the batch algorithm.
Surprisingly, the optimal set algorithm cuts time cost by several orders of magnitude, and runs in milliseconds for small graphs.
All implementations are sensitive to density, performing better when density is low.

@@

\begin{lemma}
\label{one_occurence_lemma}
A path that involves the new edge more than once is equal to a path without multiple occurrences of the edge, under the assumption that the pairs in C are verified.
\end{lemma}
\begin{proof}
Consider the part of the path between the first occurrence of the new edge and the second. This forms a cycle, which has a corresponding check in $C$ and must be verified to be equal to the identity.
This means that the loops containing the multiple occurrences of the new edge may be ignored and the path with a single occurrence of the new edge with the cycles removed is equal to this path.
\end{proof}
